---
title: "Assignment 4"
author: "Deanna Elliott"
date: "11/6/2021"
output: html_document
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

options(digits = 4) # number of per value in tables

# attach packages

library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
library(lubridate)
library(effsize)
library(broom)
```

## **Overview**

This report provides an exploratory overview of differences in juvenile snowshoe hare (*Lepus americanus*) sizes for three locations (black spruce stand, mature forest, and riparian flood zone) observed within the Bananza Creek Long Term Ecological Research (LTER) network site from 1999 - 2012 (Kielland et al. 2017). Specifically, weights are compared between sexes, and the relationship between hind foot length and body mass is explored.


## **Data & Analysis**

Snowshoe hare (*Lepus americanus*) size measurements were collected and made available by Dr. Knut Kielland and colleagues at the Bonanza Creek LTER network site in Fairbanks, Alaska. The data contain observations for 378 juvenile hares, collected over 15 years (1999 - 2012). Following exploratory data visualization, weights between male and female hares are analyzed by two-sample t-tests using a significance level (α) of 0.05 throughout. Differences between groups are described by Cohen’s d effect size. The relationship between hind foot length and body mass is explored by simple linear regression. All analyses are in R version 4.1.1 using RStudio version 1.4.1717.


## **Juvenile Hare Initial Investigation**

### Annual Juvenile Hare Trap Counts

```{r, include = FALSE}

juv_hares <- read_csv(here("data", "bonanza_hares.csv")) %>% 
  filter(age == "j") %>% 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date))        # get juvenile hare data and add year column

ann_traps <- juv_hares %>% 
  group_by(year) %>% 
  summarize(sample = n())     # find number of traps per year

trap_sum <- ann_traps %>% 
  summarize(min = min(sample), max = max(sample), 
            median = median(sample), mean = mean(sample))   # summary stats for annual traps

days_spent <- juv_hares %>% 
  group_by(session_id) %>% 
  summarize(sample = n())       # find days spent collecting trap per year
```

```{r}

# plot number of juvenile trap counts per year

ggplot(data = ann_traps,
       aes(x = year, y = sample)) +
  geom_col(fill = "cadetblue2", color = "black") +
  labs(x = "Year", y = "Sample Size") +
  theme_dark() +
  scale_x_continuous(breaks = seq(from = 1999, to = 2012, by = 2))

```

**Figure 1. Juvenile snowshoe hare trap count by year.** Over the course of 15 years (1998-2012), the Bonanza Creek Long Term Ecological Research network site collected trap observations of snowshoe hares. The minimum trap count of juvenile hares per year was `r min(ann_traps$sample)`, occurring in 2010, while the maximum trap count was `r max(ann_traps$sample)`, occurring in 1999. The mean trap count value over the course of the study was `r round(trap_sum$mean, 0)`, while the median was `r round(trap_sum$median, 0)`. After 1999, there was a sharp drop off in juvenile count by almost half, and the numbers never recovered across the life of the available data. There were spikes in 2005 and 2008, but even those did not reach the value of 2000, let alone 1999. Based on this data, it seems the count of trapped juvenile hares is fairly low annually in comparison to 1999.

.

##### **Suggestions**
These are absolute counts, but these counts are also impacted by effort. The number of days spent observing each year swings between `r min(days_spent$sample)` and `r max(days_spent$sample)`. Moving forward, I would suggest that the study regulate the number of observers and the number of days spent observing per year in order to standardize the juvenile hare population count in future analyses.


### Visualize Juvenile Hare Weights

```{r}

site_hares <-juv_hares %>% 
  mutate(grid_full = case_when(
    grid == "bonrip" ~ "Riparian",
    grid == "bonmat" ~ "Mature",
    grid == "bonbs" ~ "Black Spruce")) %>% 
  mutate(sex = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",
    is.na(sex) == TRUE ~ "Unknown"))   # reformat site names and sexes to look better in plot

ggplot(data = site_hares,
       aes(x = sex, y = weight, fill= sex)) +
  scale_fill_manual(values = c("violetred", "royalblue1", "seagreen2")) +
  geom_jitter(color = "black", pch = 21, size = 2,
              na.rm = FALSE) +
  facet_wrap(~ grid_full) +
  labs(x = "Sex", y = "Weight (g)") +
  theme_bw() +
  theme(legend.position = "none")   # plot sex vs weight for each site
```

**Figure 2. Juvenile snowshoe hare weights by sex and site.** Three sites in the Bonanza Creek LTER were observed for juvenile snowshoe hares. The riparian site yielded the highest number of both male and female hares, followed by the mature site and the black spruce site. Weights of hares at each site seemingly follow a similar distribution pattern, with male weights appearing to be slightly higher than female weights. The riparian site had a small number of weight observations above the maximums of mature and black spruce, mostly male. The few observed animals of unknown sex fall into these general distributions, excepting three at the riparian site that fall under the otherwise smallest weights. Data: Kielland et al. (2017).


## **Juvenile Weight Comparison**

### Male & Female Hares

```{r}

# find summary stats for male and female hares 

site_hares %>% 
  group_by(sex) %>%
  filter(sex %in% c("Male", "Female")) %>% 
  summarize(mean = mean(weight, na.rm = TRUE), stdev = sd(weight, na.rm = TRUE), sample = n()) %>% 
# create pretty table for summary stats by sex
  
  kbl(col.names = c("Sex", "Mean Weight (g)", "Standard Devation (g)", "Sample Size"), caption = "<b> Table 1. Summary statistics for snowshoe hare weight.</b> Mean and standard deviation in g, with sample sizes of male and female juvenile snowshoe hares observed at the Bonanza Creek LTER.") %>% 
  kable_classic()
```
 
```{r, exploration of assumptions, include = FALSE}

weight_male <- site_hares %>%    # new dataset for male weights
  filter(sex == "Male") %>% 
  drop_na(weight) %>% 
  pull(weight)

weight_female <- site_hares %>%      # new data set for female weights
  filter(sex == "Female") %>% 
  drop_na(weight) %>% 
  pull(weight)

# explore normality of data with histograms and qq plots

hist(weight_female) #pretty normal      
hist(weight_male) #normal-ish

ggplot(data = site_hares, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~ sex)

# female data looks pretty normal, male data looks normal enough
# I can use a t-test
```

```{r}

# test for difference in means

t_test <- t.test(weight_female, weight_male)

# test for effect size

cd <- cohen.d(weight_female, weight_male)
```

.

The mean weight of male snowshoe hares at Bonanza LTER is `r round(t_test$estimate[2] - t_test$estimate[1], 1)` g greater than the mean female weight; this is a 10.05% difference. If these samples were taken from populations with the same mean, there is a `r round(t_test$p.value*100, 2)`% chance of finding sample means *at least* as different as these, which is very remote. Additionally, the means of male and female weights do differ significantly (t-test | t(`r round(t_test$parameter, 0)`) = `r round(t_test$statistic, 2)`, p = `r round(t_test$p.value, 4)`). However, the effect size of the difference is small (Cohen's d |  d = `r round(cd$estimate, 2)`).


## **Relationship: Juvenile Weight & Hind Foot Length**

```{r}

# plot linear regression for weight & hind foot length

ggplot(data = juv_hares, 
       aes(y = weight, x = hindft)) +
  geom_jitter(aes(fill = age), 
              color = "black", pch = 22, size =2) +    #make square points with black outlines
  geom_smooth(method = "lm") +
  scale_fill_manual(values = "darkblue") + # change point color
  theme(legend.position = "none",
        panel.background = element_rect(fill = "cadetblue2", colour = "cadetblue",
                                size = 2, linetype = "solid"),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                colour = "white"), 
        panel.grid.minor = element_line(size = 0.25, linetype = 'solid',
                                colour = "white")) +    #change background/grid colors
  labs(x = "Hindfoot Length (mm)", y = "Weight (g)")
```

```{r, include = FALSE}

hares_lm <- lm(weight ~ hindft, data = juv_hares) #create linear model

# Coef = -279.34 (intercept), 9.52 (hindft = slope)

hares_lm_tidy <- tidy(hares_lm) #get lm stats

hares_lm_glance <- glance(hares_lm) #get model fit & stat significance
# r-sq = 0.2988, adj r-sq = 0.2959
# p-value = 8.458e-21, reject null

pr<- cor.test(juv_hares$hindft, juv_hares$weight) # test for correlation

plot(hares_lm) # look at residual plots to determine if lm assumptions are met
```

**Figure 3. Juvenile snowshoe hare hind foot length vs weight.** Measurements of juvenile hares in the Bonanza Creek LTER show a moderately positive correlation between hind foot length (in mm) and weight (in g). Generally, as the length of the hind foot increases, so does the animal's weight. Linear model summary: β1 = `r round(hares_lm$coefficients [2], 2)` g mm-1, p < 0.001, R2 = `r round(hares_lm_glance$adj.r.squared, 3)`, Pearson’s r = `r round(pr$estimate, 2)`). Data: Kielland et al. (2017).

.

The linear regression model of juvenile hare hind foot length versus weight shows a slope of `r round(hares_lm$coefficients [2], 2)` g/mm, meaning that, on average, we would expect to see a `r round(hares_lm$coefficients [2], 2)` g increase in weight for every 1 mm increase in hind foot length. The adjusted R-squared value shows that `r round(hares_lm_glance$adj.r.squared, 3)*100`% of the variance in weight is explained by hind foot length. The Pearson's r test showed an association estimate of `r round(pr$estimate, 2)`, meaning there is a moderate positive correlation between our two variables. 

The residuals for this model are fairly normal, enough that I would say it meets the assumption of normality required for linear regression, and there is no reason to suspect serial correlation in error. However, based on exploratory graphs of residuals versus fitted values, the residuals are not homoscedastic, which is an important assumption. This means our predictions for uncertainty in this model may be over- or underestimated. Additionally, the data spread is only vaguely linear. This makes me uncertain that the linear model is a good descriptor of the relationship between hind foot length and weight in juvenile hares at Bonanza Creek LTER.


## **Summary**

Exploratory data analysis reveals the following initial findings:

- Juvenile snowshoe hare population numbers seem to have fallen since 1999, though this could be due to inconsistency in trapping efforts per year.
- The riparian site yielded the highest number of trapped juveniles, both male and female.
- Male hare weight was higher than female hare weight; The difference between groups was significant, but the effect size was small.
- Hind foot length appears moderately positively correlated with body mass in juvenile hares; The relationship (by simple linear regression) is significant, with a slope (expected average increase in weight with a 1 mm increase in hind foot length) of `r round(hares_lm$coefficients [2], 2)` g/mm.

#### **Next Steps**:

- Compare hind foot lengths statistically (one-way ANOVA), including juvenile hare sex as a covariate
- Multiple linear regression to understand the relationship between sex, hind foot length, and weight


## **Citations**

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14 


.